|

### 2021-09-16 paper.Rmd
writing

---
title: |
  | Emotional framing and the effectiveness of 
  | corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the Fall 2020 CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study. This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]
author: |
  | *Working paper*
  |
  | Pirmin Stöckle^[Corresponding address: pirmin.stoeckle@gess.uni-mannheim.de]
  | GESS and SFB 884, University of Mannheim
date: \today
linestretch: 1.5
colorlinks: true
abstract: \setstretch{1} \noindent Concerns about various forms of misinformation and its fast dissemination through online media have generated huge interest into ways to effectively correct false claims. An under-explored mechanism in this research is the role of distinct emotions. How do emotional appeals interact with corrective information? Specifically, I focus on the emotion of disgust, which has been shown to be linked to the moralization of attitudes, which in turn reduces the impact of empirical evidence on attitudes and makes compromise less likely. Substantively, I investigate the issue of genetically modified (GM) food. I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000). Results show what we have learned about the question. This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims. In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations. \vspace{.8cm}
# abstract-template: Everyone agrees that this issue is really important. But we do not know much about this specific question, although it matters a great deal, for these reasons. We approach the problem from this perspective. Our research design focuses on these cases and relies on these data, which we analyze using this method. Results show what we have learned about the question. They have these broader implications.

@@

<!-- % still, we know little about the role of discrete emotions in interacting with misinformation and corrective information -->
While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2021happiness; @weeks2015emotions]. This is unfortunate given the important role that emotional appeals play in the dissemination of false or misleading information. We know that false news spread faster than true news, possibly because they are more likely to evoke emotions of surprise and disgust [@vosoughi2018spread]. In general, moral-emotional language is rewarded on social media [@brady2017emotion] and populist parties use more negative emotional appeals in their communication than non-populist parties [@widmann2021how].

@@

<!-- % get more concrete linking belief echoes / affective perseverance and specific emotion with possible connection to moralization -->
In his study, Clifford shows that disgust-evoking messages can increase moral conviction with its consequences on polarizing attitudes. However, some of the treatment material used in his study comes from a later retracted study linking GMO and cancer. Such a piece of information can be classified as misinformation in the sense of going against ``best available evidence in the public sphere''[@flynn2017nature, p.128].
From findings of belief echoes [@thorson2016belief] and affective perseverance [@sherman2002affective] we might suspect that such emotional effects of frames may persist even in the face of successful factual correction.

@@

* Definition and distinction between misinformation and disinformation.
* Some worries about backfire effects, but usually, corrective information works, and guidelines for best practice corrections have been collected.
* Emotions may contribute to the spread of false information.
* Relatively little evidence on the role of emotions for misinformation and corrective information.

<!-- % some definitions from Kuklinski and Flynn to Jerit -->

<!-- % information corrects factual beliefs, but the effects of information on attitudes are mixed -->

<!-- Factual beliefs are affected in the direction of presented evidence for both (a) misinformation and (b) corrective information. -->

<!-- The effects on related opinions are much less clear. -->
<!-- Some studies find effects, some not. -->

<!-- Affective responses are at least implicitly discussed in this literature, but discrete emotions are widely neglected. -->

@@

* Emotions feature prominently in political communication.
* Framing.
* Some known effects on information processing.
* Findings from moral psychology on moralization through disgust.
* Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition.

@@

<!-- Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.  -->

<!-- Research has shown that... anger... anxiety... enthusiasms... -->

<!-- A recently growing interest is in disgust. -->
<!-- Connection to morality, relevant for immigration, homeless ... -->

<!-- A study has shown that moral shock works to moralize attitudes towards GM food. -->

<!-- However, the main part of the material used there can be classified as misinformation -->
<!-- GM food is an issue on which disinformation exists. This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence. Least knowledgeable people are the fiercest opponents. So -->

@@

Together, hypotheses 2a and 2b represent the existence of a ``belief echo'', i.e. the continued effect of a piece of misinformation on attitudinal or emotional outcomes, despite successful factual correction. If only H2a is supported, the correction is altogether successful.

@@

Corrective information such as fact-checking is usually deliberately presented in a rather neutral way. In the case of strongly emotional framed misinformation, such a neutral correction may thus be successful in correcting factual beliefs, but not be able to counter the emotional impact of the misinformation. If this is the case, a more effective strategy may be to counter the emotional misinformation not with a neutral correction but with a correction embedded in an emotional counter-frame. If equally persuasive, such counterframes have been shown to negate the effect of an original frame [@chong2013counterframing]. I expect a similar effect to hold in response to emotionally framed misinformation.

* H3 A correction embedded in a persuasive emotional counter-frame eliminates the misinformation effect on policy opinions, moral conviction, and emotions.

@@

<!-- % explain the relevance of GMO as an issue, also in realtion to moralization -->
The topic of GM food is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does, @hasell2020differential, @carnahan2019processing]. This is especially true within the German population, which is where this study draws its sample from [@gaskell2010europeans]. GM food is also a topic on which there are misperceptions, possibly due to disinformation. Of course, there is legitimate criticism on the use of genetic modification. However, studies show that extreme opponents of GM food know the least about it [@fernbach2019extreme], and a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system [@siegrist2020consumer].

@@

To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[Pre-registration can be found on OSF: https://osf.io/y63xu.] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting]. The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany. The survey is conducted online, but recruitment was conducted offline according to a strict random probability sample. Households that did not posses the technical devices to participate online have been equipped with such devices. Fieldwork for this study took place in May 2021 and the sample size reached 4,260 respondents.

The between-subjects survey experiment followed a 3 $\times$ 3 full factorial design varying both misinformation (none, low emotional framing, high emotional framing) and corrective information (none, neutral correction, counterframe correction). Respondents were randomly assigned to one of the nine resulting groups, thus yielding eight treatment conditions and one control condition were no information treatment was shown.

An extensive debriefing is included at the end of the study for all respondents. Specifically, all original sources of information are revealed.
Because all shown information has in a similar format been in the media, there is no deception taking place.
The experimental treatment intends to evoke emotions, but I consider them within reason. There are two main reasons for that. First, the material including the used imagery has been used in a previously published study [@clifford2019emotional]. Second, the central part of the visual material has been used by actual reporting on the issue in mainstream media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt].
The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is an appropriately pleasant experience for respondents. Because this approval has been granted, this study did not require further approval by an ethics committee.

@@

The experimental treatment consists of a combination of two types of information. One type of information reports the results of a study claiming that genetically modified corn lead to increased cancer rates in rats. The wording is partly taken from actual news coverage of the study in 2012, and partly from treatment material in @clifford2019emotional. The respective study has been retracted due to non-adherence to scientific procedures, and the current scientific evidence does not point to severe health problems associated with genetically modified animal feed. The information within the study can thus be classified as misinformation.^[This is consistent with a definition of resulting mis*perceptions* as "factual beliefs that are false or contradict the best available evidence in the public domain" [@flynn2017nature, p.128]. If one would presume the intent to mislead, this would then be classified as disinformation.] I vary this misinformation to be either low or high in emotional framing by ex- or including graphical images of a rat with a cancer tumor and rotten corn, images that are taken from @clifford2019emotional and shown to induce negative feelings, specifically disgust.

The second type of information reports the accurate scientific consensus that GM food is generally safe to eat, that a study showing cancer growth in rats due to GM feed has been retracted, and that genetically modifying plants for the food industry is not a new phenomenon. This piece of information is a correction to the misinformation in that it reflects best available scientific evidence and it is constructed to conform to best practice standards of debunking information [@lewandowsky2020debunking]. As a neutral correction, this piece of information corrects the misinformation about concrete health risks from GM food in a text-only format. By negating the risks, it still stays within the overall framing of GM food as possibly risky, and does not add specific further information about benefits from GM food. This is done in a further condition, where the corrective information is embedded in a counter-frame, emphasizing positive aspects of GM food, such as its potential to contribute to more sustainable agriculture by reducing pesticides, and its potential to fight malnutrition.

Assignment is done by simple random assignment into equal sized groups. Table gives an overview over the nine resulting experimental groups. The misinformation (if any is assigned) is always shown first, immediately followed by the corrective information (if any is assigned). For respondents, the manipulation is thus designed in a way that both types of information ``hits them between the eyes'' [@kuklinski2000misinformation, p.805].

@@

As the factual belief of interest, I measure factual beliefs about the safety of GM foods by asking respondents whether they think that the statement that eating GM foods is safe for one's health is false or correct on a 7-point scale from "certainly false" to "certainly correct".

To provide a broad measure of policy support, I use a measure by [@diamond2020does] that asks respondents how high a policy priority it should be to restrict the use of GM food on a 7-point scale from "no priority" to "highest priority".

I measure moral conviction with two standard items from the literature, asking respondents whether their evaluation of GM food is connected to moral convictions and about fundamental right and wrong, respectively, on a 5-point scale from "not at all" to "very much" [@skitka2020psychology]. This is a direct measure of a "meta-cognition", building on the assumption that respondents know which issues are about moral conviction for them. Both items will be combined into a simple additive index.

I measure distinct emotions by directly asking respondents how well a given feeling represents how they feel toward GM food. Asking about their feelings toward GM food makes this a measure of integral, not incidental emotion, in line with expectation about to which kind of emotion moralization responds [@skitka2018attitude]. The distinct negative emotions I measure are disgust, anger, anxiety, and sadness, each on a 5-point scale from "not at all" to "extremely".

@@

First, \@ref(fig:misinfo-only) reports the effects of reiceiving only the misinformation message. The upper half of the figure shows two plots with the effects of the low emotion misinformation and the lower half shows two plots with the effects of the high emtoion misinformation message. The y-axis indicates the different outomce variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.

```{r misinfo-only, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
# H1: misinformation effect
pdata_h1_belief_policy <- rbind(get_pdata(lm_moral_conviction, c(4, 7), "Moral conviction"),
                                get_pdata(lm_policy_opinion, c(4, 7), "Policy restrictions"),
                                get_pdata(lm_knowledge_safe, c(4, 7), "Belief GMO safety"))

@@

```{r fig-2, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

@@

```{r fig-3, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

@@

* Persuasive, emotionally framed misinformation can increase moralization of opinions about GM food.
* Effects persist with neutral corrective information, but counter-framing works.
* Overall small effects: Explained variance ($R^2$) of $7 \%$ for beliefs, $3.5 \%$ for opinions, $1 \%$ for moral conviction.

@@

* No distinction between emotional framing and visual cues.

@@

* When misinformation is emotionally charged, fact-checkers may have to think about counterframes to eliminate the misinformation effect.
* a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system @siegrist2020consumer.  genetic engineering provides avenues with large potential benefits @roberts2018nobel, @planck2019scientists, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
* Similar effects possible on other issues where misinformation loads on strong emotions, e.g. health interventions such as vaccinations.

@@

## Information effect

```{r fig-4, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

>>>>

---
title: |
  | Emotional framing and the effectiveness of 
  | corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the Fall 2020 CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study. This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on "The Political Economy of Reforms" (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]
author: |
  | *Working paper*
  |
  | Pirmin Stöckle^[Corresponding address: pirmin.stoeckle@gess.uni-mannheim.de]
  | GESS and SFB 884, University of Mannheim
date: \today
linestretch: 1.5
colorlinks: true
abstract: \setstretch{1} \noindent Concerns about various forms of misinformation and its fast dissemination through online media have generated huge interest into ways to effectively correct false claims. An under-explored mechanism in this research is the role of distinct emotions. How do emotional appeals interact with corrective information? Specifically, I focus on the emotion of disgust, which has been shown to be linked to the moralization of attitudes, which in turn reduces the impact of empirical evidence on attitudes and makes compromise less likely. Substantively, I investigate the issue of genetically modified (GM) food. I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000). Results show what we have learned about the question. This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims. In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations. \vspace{.8cm}
# abstract-template: Everyone agrees that this issue is really important. But we do not know much about this specific question, although it matters a great deal, for these reasons. We approach the problem from this perspective. Our research design focuses on these cases and relies on these data, which we analyze using this method. Results show what we have learned about the question. They have these broader implications.

@@

<!-- % still, we know little about the role of discrete emotions in interacting with misinformation and corrective information -->
While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored. This is unfortunate given the important role that emotional appeals play in the dissemination of false or misleading information. We know that false news spread faster than true news, possibly because they are more likely to evoke emotions of surprise and disgust [@vosoughi2018spread]. In general, moral-emotional language is rewarded on social media [@brady2017emotion].
<!-- and populist parties use more negative emotional appeals in their communication than non-populist parties [@widmann2021how]. -->

@@

<!-- % get more concrete linking belief echoes / affective perseverance and specific emotion with possible connection to moralization -->
In his study, Clifford shows that disgust-evoking messages can increase moral conviction with its consequences on polarizing attitudes. However, some of the treatment material used in his study comes from a later retracted study linking GMO and cancer. Such a piece of information can be classified as misinformation in the sense of going against "best available evidence in the public sphere"[@flynn2017nature, p.128].
From findings of belief echoes [@thorson2016belief] and affective perseverance [@sherman2002affective] we might suspect that such emotional effects of frames may persist even in the face of successful factual correction.

@@

* Definition of related terms: misinformation, disinformation, fake news, misperceptions [@flynn2017nature; @jerit2020political; @kuklinski2000misinformation].
* Wide-spread misperceptions may distort public opinion and corrective information (fact-checking) is one way to tackle the problem.
* People may resist corrective information due to directional motivated reasoning, sometimes even leading to "backfire" effects [@nyhan2010corrections] where misperceptions are more strongly held after correction, but most of the time corrections are at least partially successful [@walter2020fact; @wood2019elusive].
* However, while corrective information usually moves factual beliefs in the appropriate direction, the effects on other outcomes such as attitudes, blame attributions, or evaluations are mixed [@bisgaard2019getting; @nyhan2019taking; @swire2019they]
* An especially troubling case are "belief echoes", where effects of misinformation on attitudes persist despite successful correction of the factual belief [@thorson2016belief].
* Although we know that emotions may contribute to the spread of false information [@vosoughi2018spread], there is little evidence on the role of emotions for misinformation and corrective information (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2021happiness; @weeks2015emotions]).

@@

* Emotional framing is an important feature of political communication, and theories such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.
* Findings from moral psychology on moralization through disgust.
* Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition.
* Of course, people are free to have moral convictions about any issue, but if emotional frames can induce moral conviction, we might have to worry about emotionally framed **mis**information also inducing moral conviction.

@@

@@

Together, hypotheses 2a and 2b represent the existence of a "belief echo", i.e. the continued effect of a piece of misinformation on attitudinal or emotional outcomes, despite successful factual correction. If only H2a is supported, the correction is altogether successful.

@@

Corrective information such as fact-checking is usually deliberately presented in a rather neutral way. In the case of strongly emotional framed misinformation, such a neutral correction may thus be successful in correcting factual beliefs, but not be able to counter the emotional impact of the misinformation. If this is the case, a more effective strategy may be to counter the emotional misinformation not with a neutral correction but with a correction embedded in an emotional counter-frame. If equally persuasive, such counterframes have been shown to negate the effect of an original frame [@chong2007competitive; @chong2013counterframing]. I expect a similar effect to hold in response to emotionally framed misinformation.

* H3 A correction embedded in a persuasive emotional counterframe eliminates the misinformation effect on policy opinions, moral conviction, and emotions.

@@

<!-- % explain the relevance of GMO as an issue, also in relation to moralization -->
The topic of GM food is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does; @hasell2020differential; @carnahan2019processing]. This is especially true within the German population, which is where this study draws its sample from [@gaskell2010europeans]. GM food is also a topic on which there are misperceptions, possibly due to disinformation. Of course, there is legitimate criticism on the use of genetic modification. However, studies show that extreme opponents of GM food know the least about it [@fernbach2019extreme], and a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system [@siegrist2020consumer].

@@

To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[The pre-registration can be found on OSF: https://osf.io/y63xu.] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting]. The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany. Fieldwork for this study took place in May 2021 and the sample size reached 4,260 respondents.

The between-subjects survey experiment followed a 3 $\times$ 3 design varying both misinformation (none, low emotional framing, high emotional framing) and corrective information (none, neutral correction, counterframe correction). Respondents were randomly assigned to one of the nine resulting groups, thus yielding eight treatment conditions and one control condition were no information treatment was shown.

At the end of the survey, respondents received an extensive debriefing clarifying the sources of all information that was shown and on the intention of the experiment. Because all shown information has in a similar format been in the media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt], there is no deception taking place. The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is a pleasant experience for respondents.

@@

The experimental treatment consists of a combination of two types of information, each mimicking a short online news article. One type of information reports the results of a study claiming that genetically modified corn lead to increased cancer rates in rats, and features the headline *"GM corn-fed rats die earlier"*. The wording is partly taken from actual news coverage of the study in 2012, and partly from treatment material in @clifford2019emotional. The respective study has been retracted due to non-adherence to scientific procedures, and the current scientific evidence does not point to severe health problems associated with genetically modified animal feed [@brody2018are]. The information within the study can thus be classified as misinformation.^[This is consistent with a definition of resulting mis*perceptions* as "factual beliefs that are false or contradict the best available evidence in the public domain" [@flynn2017nature, p.128]. If one would presume the intent to mislead, this would then be classified as disinformation.] I vary this misinformation to be either low or high in emotional framing by ex- or including graphical images of a rat with a cancer tumor and rotten corn, images that are taken from @clifford2019emotional and shown to induce negative feelings, specifically disgust.

The second type of information reports the accurate scientific consensus that GM food is generally safe to eat, that a study showing cancer growth in rats due to GM feed has been retracted, and that genetically modifying plants for the food industry is not a new phenomenon. This piece of information is a correction to the misinformation in that it reflects best available scientific evidence and it is constructed to conform to best practice standards of debunking information [@lewandowsky2020debunking]. As a neutral correction, this piece of information corrects the misinformation about concrete health risks from GM food in a text-only format. The headline for the neutral correction reads *"Genetic engineering does not introduce novel risks"*. By negating the risks, it still stays within the overall framing of GM food as possibly risky, and does not add specific further information about benefits from GM food. This is done in a further condition, where the corrective information is embedded in a counterframe, emphasizing positive aspects of GM food, such as its potential to contribute to more sustainable agriculture by reducing pesticides, and its potential to fight malnutrition. Here, the headline reads *"Genetic research for a more sustainable agriculture"*. ^[Screenshots of the shown information treatments in the German original are included in the Appendix in figures \@ref(fig:screenshot-low-misinfo) to \@ref(fig:screenshot-counterframe).]

The misinformation (if any is assigned) is always shown first, immediately followed by the corrective information (if any is assigned). For respondents, the manipulation is thus designed in a way that both types of information "hits them between the eyes" [@kuklinski2000misinformation, p.805].

@@

To acknowledge findings from the maturing corrective information literature on varying correction success for different outcomes, I include a set of several measures.

As the factual belief of interest, I measure factual beliefs about the safety of GM foods by asking respondents whether they think that the statement that eating GM foods is safe for one's health is false or correct on a 7-point scale from "certainly false" to "certainly correct". To provide a broad measure of policy support, I use a measure by [@diamond2020does] that asks respondents how high a policy priority it should be to restrict the use of GM food on a 7-point scale from "no priority" to "highest priority". I measure moral conviction with two standard items from the literature, asking respondents whether their evaluation of GM food is connected to moral convictions and about fundamental right and wrong, respectively, on a 5-point scale from "not at all" to "very much" [@skitka2020psychology]. This is a direct measure of a "meta-cognition", building on the assumption that respondents know which issues are about moral conviction for them. Both items will be combined into a simple additive index.

Lastly, I measure distinct emotions by directly asking respondents how well a given feeling represents how they feel toward GM food. Asking about their feelings toward GM food makes this a measure of integral, not incidental emotion, in line with expectation about to which kind of emotion moralization responds [@skitka2018attitude]. The distinct negative emotions I measure are disgust, anger, anxiety, and sadness, each on a 5-point scale from "not at all" to "extremely".

@@

First, \@ref(fig:misinfo-only) reports the effects of receiving only the misinformation message. The upper half of the figure shows two plots with the effects of the low emotion misinformation and the lower half shows two plots with the effects of the high emtoion misinformation message. The y-axis indicates the different outcome variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.

In the top row of the lower right panel, we see the largest effect, which results from the high emotional frame on the factual belief outcome. Respondents receiving the high emotional misinformation about GM food agree to the statement that GM food is safe to eat with a score that is 16 percentage points less than the control group. For the low emotion misinformation, the effect is 13 percentage points, indicating that both messages are persuasive in bringing the average factual belief farther from the scientific consensus belief.

Concerning effects on emotions, policy opinion and moral conviction, results largely replicate results by @clifford2019emotional.

```{r misinfo-only, fig.align="center", fig.cap="Effects of misinformation with low and high emotional framing without following correction. Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
# H1: misinformation effect
pdata_h1_belief_policy <- rbind(get_pdata(lm_moral_conviction, c(4, 7), "Moral conviction"),
                                get_pdata(lm_policy_opinion, c(4, 7), "Policy restrictions"),
                                get_pdata(lm_knowledge_safe, c(4, 7), "Belief GMO safety"))

@@

After having established that a persuasive, emotionally framed message containing misinformation can influence policy opinion and increase moral conviction, I now report the results of the neutral correction in figure \@ref(fig:neutral-correction).

* the neutral correction is effective for the low emotion misinformation, but not the high emotion misinformation

```{r neutral-correction, fig.align="center", fig.cap="Effects of the neutral correction for misinformation with low and high emotional framing Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

@@

Finally, results from the counterframe message show that a correction embedded in a persuasive positive frame is successful in bringing all outcome measures to the control group level (see figure \@ref(fig:counterframe). 
It should be noted, however, that there is also no positive effect of the positively framed message on beliefs and opinions towards GM food, although the counterframe message itself is persuasive (figure \@ref(fig:info-only) in the appendix).

```{r counterframe, fig.align="center", fig.cap="Effects of the counterframe correction for misinformation with low and high emotional framing. Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

@@

* Persuasive, emotionally framed misinformation can increase moralization of opinions about GM food, replicating findings from @clifford2019emotional.
* A neutral correction is successful for low emotional misinformation, but effects of highly emotional misinformation persist with neutral corrective information
* Providing a correction embedded in a positive counterframe message works for both low and high emotional misinformation.
* Overall small effects: Explained variance ($R^2$) of $7 \%$ for beliefs, $3.5 \%$ for opinions, $1 \%$ for moral conviction.

@@

* No distinction between emotional framing and visual cues (see e.g. @hameleers2020picture).

@@

* When misinformation is emotionally charged, fact-checkers and science communicators may have to think about counterframes to eliminate the misinformation effect.
* Concerning GM food, it is argued that a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system [@siegrist2020consumer]. With emotional misinformation in circulation, persuasive counterframes seem necessary to even equal out their effects.
* Similar effects seem likely on other issues where misinformation loads on strong emotions, e.g. health interventions such as vaccinations.

@@

## Information effects

```{r info-only, fig.align="center", fig.cap="Effects of the neutral correction and counterframe correction without previous misinformation. Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
